Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-21 paper.tex
todos
cite ron for the state of art of threesomes
cite the right garcia for supercoercions
sec 3 in much more details

\newcommand{\rOOsucc}[1]{#1}
\newcommand{\rOOfail}[1]{#1}
%\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
%\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

Around 2006, several groups of researchers proposed ways to integrate
dynamic typing and static typing, notably gradual typing
\citep{siek2006gradual}, hybrid typing \citep{flanagan2006hybrid},
migratory typing \citep{Tobin-Hochstadt:2006fk}, and multi-language
iteroperability \citep{Gray:2005ij,Matthews:2007zr}. Researchers
usually define the semantics of gradually typed languages by
translation to an intermediate language with casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi
\citep{siek2009exploring}. Unfortunately, straightforward
implementations of casts on higher-order values (functions, objects,
etc.) impose significant runtime overheads that can change the
asymptotic time and space complexity of a program
\citep{herman2010space}. There are several known space-efficient cast
representations, with various strengths and weaknesses
\citep{siek2015blame,siek2010threesomes,garcia2013calculating,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}.
The current state of the art includes
\begin{itemize}
\item threesomes \citep{siek2010threesomes},
\item supercoercions \citep{garcia2014deriving}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
Recall that in these systems, casts are compressed using a composition
operator.  Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

@@

This section presents our first contribution, the definition of hypercoercion. 
The design of hypercoercion is motivated by observations on coercion normal 
forms: a normal coercion has at most three parts; projections are always at the 
beginning; injections and failures are always at the end. Hypercoercions have 
similar shape: a hypercion has either one part or three parts; for three-part 
hypercoercions, the first part is either a projection or a no-op, while the 
last part is either an injection, a failure, or a no-op.

\figref{fig:hypercoercion} defines the syntax of hypercoercion. Types, 
pre-types, and blame labels are as before.

Let $ c $ range over hypercoercions. A hypercoercion is either 
an identity cast between the dynamic types, or a complex including a head, a 
middle, and a tail. 
Let $ h $ range over heads. A head is either a no-op, or a projection.
Let $ m $ range over middles. There is a one-to-one 
correspondence between middles and type constructors. 
We generalize shallow-consistency to middles $ m \smile m $ in the natural way.
Let $ t $ range over tails. A tail is either a no-op, an injection, or a 
failure.

We define hypercoercion constructs in the following subsections.
But functions that apply hypercoercions to values are deferred to 
Section~\ref{sec:hypercoercion-correctness} because 
they depends on a new definition of values, which is part of the framework in
Section~\ref{sec:framework}.

\subsection{\lazyD{} hypercoercion}

@@

	\comprule{
		\hyperCoercionC{h_1}{m_1}{t_1}
	}{
		\hyperCoercionC{\epsilon}{m_2}{t_2}
	}{
		\hyperCoercionC{h_1}{m_1 \fatsemi^{\ell} m_2}{t_2}
	}{
		\sidecond{
			m_1 \smile m_2
			\; \text{and} \;
			\forall l. t \neq \bot^{l}
		}
	}
	\comprule{
		\hyperCoercionC{h_1}{m_1}{t_1}
	}{
		\hyperCoercionC{?^l}{m_2}{t_2}
	}{
		\hyperCoercionC{h_1}{m_1 \fatsemi^{l} m_2}{t_2}
	}{
	\sidecond{
		m_1 \smile m_2
		\; \text{and} \;
		\forall l. t \neq \bot^{l}
	}
	}
	\comprulel{
	\hyperCoercionC{h_1}{m_1}{t_1}
	}{
	\hyperCoercionC{h_2}{m_2}{t_2}
	}{
	\hyperCoercionC{h_1}{m_1}{\bot^l}
	}{
	\sidecond{
		\neg \; m_1 \smile m_2
		\; \text{and} \;
		\forall l. t \neq \bot^{l}
	}
	}
%
%	\comprule{
%		\hyperCoercionC{h}{m_1}{t_1}
%	}{
%		\hyperCoercionC{\epsilon}{m_2}{t_2}
%	}{
%		\hyperCoercionC{h}{m'}{t'}
%	}{
%		\sidecond{
%			m_1 \fatsemi^{\ell} (m_2, t_2) = (m', t')
%			 \; \text{and} \;
%			 \forall l. t \neq \bot^{l}
%		}
%	}
%	
%	\comprule{
%		\hyperCoercionC{h}{m_1}{t_1}
%	}{
%		\hyperCoercionC{?^{l'}}{m_2}{t_2}
%	}{
%		\hyperCoercionC{h}{m'}{t'}
%	}{
%		\sidecond{
%			m_1 \fatsemi^{l'} (m_2, t_2) = (m', t')
%			\; \text{and} \;
%			\forall l. t \neq \bot^{l} 
%		}
%	}
	\end{array}
	\]

@@

$ c_1 \fatsemi^\ell c_2 = c' $ composes two hypercoercions. The target of $ 
c_1 $ and the source of $ c_2 $ might be different, in which case $ \ell $ must 
be a label. The last two lines are the most interesting. When the head of the 
second hypercoercion is a no-op, we use the same $ \ell $ to compose the 
middles. Otherwise, the head must be a projection, where we use the label in 
the projection to compose middles.

Let $ \ell $ range over $ \epsilon $ and labels.
$ m_1 \fatsemi^\ell (m_2,t) = (m',t') $ composes a middle with a 
pair of a middle and a tail. Similarly, the target of $ m_1 $ and the source of 
$ m_2 $ might be different, in which case $ \ell $ must be a label.
$ seq(c_1,c_2) $ composes two coercions where the target type of $ c_1 $ is 
equal to the source type of $ T_2 $.
$ id(T) $ constructs an identity coercion of $ T $ with the help of $ id(P) $.
$ cast(T_1,l,T_2) $ constructs a coercion from a source type, a label, and a 
target type.

\begin{proposition}[\lazyD\ hypercoercion is a monoid]
	\label{thm:hc-monoid}
	For all $c : T_1 \Longrightarrow T_2$,
	$c_1 : T_1 \Longrightarrow T_2$,
	$c_2 : T_2 \Longrightarrow T_3$, and
	$c_3 : T_3 \Longrightarrow T_4$,
	\begin{enumerate}
		\item $seq(id(T_1),c) = c$,
		\item $seq(c,id(T_2)) = c$, and
		\item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
	\end{enumerate}
\end{proposition}

\subsection{\lazyUD{} hypercoercion [Jeremy]}

@@

\begin{lemma}[$ \text{Lazy D Cast ADT} \cap \text{Monoid} $ Respects 
	$ \mathcal{C} $]
	\label{thm:surely-monoidic-eval}
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $ and $ C $ is a 
	\lazyD{} Cast ADT and a monoid,
	\[
	\text{\evalEqv{\effCEK{C}}{\ineffCEKD}}
	\]
\end{lemma}
\begin{proof}Proving left implies right is trivial.
To prove $ \judgeDeval{e}{o} $ implies $ \judgeSeval{C}{e}{o} $, 
we firstly take one steps in \effCEK{C}, and catch it up in \ineffCEKD\ by 
taking one or more steps in \ineffCEKD. As we just took a positive number of 
steps, we are closer to the halting state.
\end{proof}

>>>>

%\newcommand{\rOOsucc}[1]{#1}
%\newcommand{\rOOfail}[1]{#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

Around 2006, several groups of researchers proposed ways to integrate
dynamic typing and static typing, notably gradual typing
\citep{siek2006gradual}, hybrid typing \citep{flanagan2006hybrid},
migratory typing \citep{Tobin-Hochstadt:2006fk}, and multi-language
iteroperability \citep{Gray:2005ij,Matthews:2007zr}. Researchers
usually define the semantics of gradually typed languages by
translation to an intermediate language with casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi
\citep{siek2009exploring}. Unfortunately, straightforward
implementations of casts on higher-order values (functions, objects,
etc.) impose significant runtime overheads that can change the
asymptotic time and space complexity of a program
\citep{herman2010space}. There are several known space-efficient cast
representations, with various strengths and weaknesses
\citep{siek2015blame,siek2010threesomes,garcia2013calculating,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}.
The current state of the art includes

\begin{itemize}
\item threesomes \citep{siek2010threesomes,garcia2013calculating},
\item supercoercions \citep{garcia2013calculating}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
Recall that in these systems, casts are compressed using a composition
operator.  Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

@@

This section presents our first contribution, the definition of hypercoercion. 
The design of hypercoercion is motivated by observations on coercion normal 
forms: a normal coercion has at most three parts; projections are always at the 
beginning; injections and failures are always at the end. Hypercoercions have 
similar shape: a hypercion either is $ \hyperCoercionI $, the identity cast for 
$ \TOOdyn $, or contains three parts, where the first part is either a 
projection or a no-op while last part is either an injection, a failure, or a 
no-op.

\figref{fig:hypercoercion} defines the syntax of hypercoercions. Types, 
pre-types, and blame labels are as before.

Let $ c $ range over hypercoercions. A hypercoercion either is the identity 
cast between the dynamic types, or includes a head, a middle, and a tail. 
Let $ h $ range over heads. A head is either a no-op, or a projection.
Let $ m $ range over middles. There is a one-to-one 
correspondence between middles and type constructors. 
We generalize shallow-consistency to middles $ m \smile m $ in the obvious way.
Let $ t $ range over tails. A tail is either a no-op, an injection, or a 
failure.

Subsection~\ref{sec:ld-hc} defines functions that construct \lazyD\ 
hypercoercions. Subsection~\ref{sec:lud-hc} defines the \lazyUD\ counterparts. 
Functions that apply hypercoercions to values are deferred to 
Section~\ref{sec:hypercoercion-correctness} because they depends on a new 
definition of values, which is part of the framework in
Section~\ref{sec:framework}.

\subsection{\lazyD{} hypercoercion}
\label{sec:ld-hc}

@@

	\comprule{
		\hyperCoercionC{h_1}{m_1}{t_1}
	}{
		\hyperCoercionC{?^l}{m_2}{t_2}
	}{
		\hyperCoercionC{h_1}{m_1}{\bot^l}
	}{
		\sidecond{
			\neg \; m_1 \smile m_2
			\; \text{and} \;
			\forall l. t \neq \bot^{l}
		}
	}

\comprulel{
\hyperCoercionC{h_1}{m_1}{t_1}
}{
\hyperCoercionC{\epsilon}{m_2}{t_2}
}{
\hyperCoercionC{h_1}{m_1}{\bot^l}
}{
\sidecond{
	\neg \; m_1 \smile m_2
	\; \text{and} \;
	\forall l. t \neq \bot^{l}
}
}
\comprule{
\hyperCoercionC{h_1}{m_1}{t_1}
}{
\hyperCoercionC{?^l}{m_2}{t_2}
}{
\hyperCoercionC{h_1}{m_1 \fatsemi^{l} m_2}{t_2}
}{
\sidecond{
	m_1 \smile m_2
	\; \text{and} \;
	\forall l. t \neq \bot^{l}
}
}
	\comprule{
		\hyperCoercionC{h_1}{m_1}{t_1}
	}{
		\hyperCoercionC{\epsilon}{m_2}{t_2}
	}{
		\hyperCoercionC{h_1}{m_1 \fatsemi^{\ell} m_2}{t_2}
	}{
		\sidecond{
			m_1 \smile m_2
			\; \text{and} \;
			\forall l. t \neq \bot^{l}
		}
	}
%
%	\comprule{
%		\hyperCoercionC{h}{m_1}{t_1}
%	}{
%		\hyperCoercionC{\epsilon}{m_2}{t_2}
%	}{
%		\hyperCoercionC{h}{m'}{t'}
%	}{
%		\sidecond{
%			m_1 \fatsemi^{\ell} (m_2, t_2) = (m', t')
%			 \; \text{and} \;
%			 \forall l. t \neq \bot^{l}
%		}
%	}
%	
%	\comprule{
%		\hyperCoercionC{h}{m_1}{t_1}
%	}{
%		\hyperCoercionC{?^{l'}}{m_2}{t_2}
%	}{
%		\hyperCoercionC{h}{m'}{t'}
%	}{
%		\sidecond{
%			m_1 \fatsemi^{l'} (m_2, t_2) = (m', t')
%			\; \text{and} \;
%			\forall l. t \neq \bot^{l} 
%		}
%	}
	\end{array}
	\]

@@

Let $ \ell $ range over $ \epsilon $ and labels.
$ c_1 \fatsemi^\ell c_2 $ composes hypercoercion $ c_1 $ and hypercoercion $ 
c_2 $.
When both $ c_1 $ and $ c_2 $ are $ \hyperCoercionI $, their 
composition is also $ \hyperCoercionI $.
When $ c_1 $ is $ \hyperCoercionI $ but $ c_2 $ is not, the head of the 
composition must be a projection. In this 
case, if the head of $ c_2 $ is a projection, we reuse it. Otherwise, we need a 
label to build the projection. Since the head of $ c_2 $ is $ \epsilon $, its 
source type must be a pretype. Thus we know $ \ell $ must be a label and put it 
in the projection. 
When $ c_1 $ ends with a failure, the composition is $ c_1 $ itself. In all 
remaining cases, we shall assume that $ c_1 $ does not end with 
a failure.
When $ m_1 $ and $ m_2 $ have different top constructors ($ \neg\;m_1 \smile 
m_2 $), proceeding to $ m_2 $ can not make sense. So we end the composition 
with a failure. To do so we need a label. When $ c_2 $ starts with a 
projection, we shall accuse it of casting a value to a shallowly inconsistent 
type. When $ c_2 $ starts with a no-op, $ \ell $ must be a label. And we accuse 
it of composing shallowly inconsistent hypercoercions. The last two cases 
compose $ m_1 $ and $ m_2 $ with a helper function $ m \fatsemi^\ell m $, which 
assumes its inputs have the same top constructor. The definition of $ m 
\fatsemi^\ell m $ is straightforward. Going back to the last two cases of $ c_1 
\fatsemi^\ell c_2 $. When $ c_2 $ starts with a projection, we have no clue 
whether the target type of $ m_1 $ is the same as the source of $ m_2 $, 
although we know they are shallowly consistent. So we give the $ l $ in the 
projection to $ m \fatsemi^\ell m $. When $ c_2 $ starts with a no-op, we reuse 
the $ \ell $.

$ seq(c_1,c_2) $ requires that the target type of $ c_1 $ is the same as the 
source of $ c_2 $. $ c \fatsemi^\ell c $ does the job for it.
$ id(T) $ constructs an identity coercion of $ T $ with the help of $ id(P) $. 
Their definitions are straightforward.
$ cast(T_1,l,T_2) $ constructs a coercion from a source type, a label, and a 
target type. $ c \fatsemi^\ell c $ does the most job.

\begin{proposition}[\lazyD\ hypercoercion is a monoid]
	\label{thm:hc-monoid}
	For all $c : T_1 \Longrightarrow T_2$,
	$c_1 : T_1 \Longrightarrow T_2$,
	$c_2 : T_2 \Longrightarrow T_3$, and
	$c_3 : T_3 \Longrightarrow T_4$,
	\begin{enumerate}
		\item $seq(id(T_1),c) = c$,
		\item $seq(c,id(T_2)) = c$, and
		\item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
	\end{enumerate}
\end{proposition}
\begin{proof}
	By induction on hypercoercion(s).
\end{proof}

\begin{proposition}[]
	For all $ T $ and $ l $, $ cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
	By induction on T.
\end{proof}

\subsection{\lazyUD{} hypercoercion [Jeremy]}
\label{sec:lud-hc}

@@

\todo[inline]{An extra property is required after changing case to small-step 
-KC}
\begin{lemma}[$ \text{Lazy D Cast ADT} \cap \text{Monoid} $ Respects 
	$ \mathcal{C} $]
	\label{thm:surely-monoidic-eval}
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $ and $ C $ is a 
	\lazyD{} Cast ADT and a monoid,
	\[
	\text{\evalEqv{\effCEK{C}}{\ineffCEKD}}
	\]
\end{lemma}
\begin{proof}Proving left implies right is trivial.
To prove $ \judgeDeval{e}{o} $ implies $ \judgeSeval{C}{e}{o} $, 
we firstly take one steps in \effCEK{C}, and catch it up in \ineffCEKD\ by 
taking one or more steps in \ineffCEKD. As we just took a positive number of 
steps, we are closer to the halting state.
\end{proof}
